```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Correlation Between COVID Cases and Vaccination Status Among Different Age Groups in the U.S.

Shelby Pankratz

Minnesota State University Moorhead Biosciences Department

## Abstract

This is an exploratory data analysis using two datasets from the CDC website. The first is comparing percentage of fully vaccinated individuals among different age groups in the U.S. over a certain period of time, and the second is comparing COVID-19 cases among different age groups in the U.S. over a certain period of time. RStudio was used to filter raw data to use specifically for this study to create tables and graphs to display certain variables. Although it was hard to relate the two datasets to each other because of the difference in age groups, a Chi-Squared test was used to compare among the different age groups in the case dataset. It was found that cases among different age groups varied greatly, as seen in Fig. 2. This was probably due to a variety of outside factors, but the 21-30 year old age group had the highest increase in COVID-19 cases in the designated time period, and the 18-24 year old age group had the lowest percentage of vaccination (not including the 5-11 year old age group, as they were not approved until later for the vaccine).

## Introduction

COVID-19 vaccinations in the U.S. first became available to select populations, such as the elderly and immunocompromised, back in June 2020, right after COVID-19 broke out in March. Fast forward to present day where vaccinations are available to everyone, along with the development of a booster shot. However, since then, vaccines have become a very controversial topic. A handful of reasons that people have been deferring vaccination are concerns about side effects and safety of the vaccine, lack of trust in the government, and the vaccine being developed too quickly (Nguyen, et al.). In the United States, it's been seen that longer testing and increased efficacy and development of a vaccine have a significant impact on vaccination acceptance (Pogue et al.). This has led to many people wondering, how effective are these vaccines?

This project will be demonstrating an exploratory data analysis on the relationship between the number of COVID-19 cases and the number of people vaccinated. This will help to show the effectiveness of the vaccine. Data will be used from COVID Data Tracker, which is a website managed by the CDC, to look for relationships between positive COVID-19 cases and vaccination status among different age groups. The whole population of the U.S. will be looked at to get more accurate information on the effectiveness of the COVID-19 vaccine. If the vaccine is effective, we should see that as the number of vaccinated people goes up, the amount of COVID-19 cases will go down.

## Methods

Data acquisition of the data sets from the CDC website is listed in the steps below.

### Data Acquisition

COVID data: Cases

1.  Go to <https://covid.cdc.gov/covid-data-tracker/#demographics>
2.  Scroll to "View Data"
3.  Click on the Export Tab
4.  Click on CSV
5.  Save to data folder (Public_Dataset_Age_1\_)

COVID data: Vaccination Status

1.  Go to <https://data.cdc.gov/Vaccinations/COVID-19-Vaccination-and-Case-Trends-by-Age-Group-/gxj9-t96f>

2.  Scroll to "View Data"

3.  Click on the Export Tab

4.  Click on CSV

5.  Save to data folder (COVID_19_Vaccination_and_Case_Trends_by_Age_Group_United_States_1\_.csv)

After downloading the data, the first step is to run the tidyverse code, then rename the data set, starting with the vaccination data set.

    library(tidyverse)

    vax_data <- read_csv("data/COVID-19_Vaccination_Demographics_in_the_United_States_National (1).csv")

To get rid of the unwanted variables and to make the variables that are to be kept a little easier to read, the rename, filter, and mutate functions are used to get the desired table. This line of code specifically works on renaming and mutating the column that deals with the age groups.

    filter(str_starts(Demographic_category, "Ages_")) %>% 
      rename(age_group = Demographic_category) %>% 
      mutate(
        Date = as.Date(Date, format = "%m/%d/%Y"),
        age_group=str_remove(age_group, "Ages_"),
        age_group=str_remove(age_group, "_yrs"),
        age_group=str_remove(age_group, "yrs")
      ) %>% 

The next step that took place was making sure that all the age groups were included in the table. Because of this, a line of code had to be added beforehand to make sure that R could read the age data

    age_group_levels <- c("<5", "5-11", "<12", "16-17", "18-24", 
                          "25-39", "40-49", "50-64", "65-74", "75+")

After this code, R is able to recognize the data set "age_group_levels" and the mutate function is used to factor the age groups.

    mutate(age_group = factor(age_group, levels = age_group_levels)) %>%

In this data set, there were a couple of age groups that overlapped with each other, or ones that weren't needed. The filter function was used to get remove them from the table as well.

    filter(!age_group %in% c("<12yrs", "<5yrs", "12-15", "16-17", "<12", "<5")) %>%

After that, rename the other columns so they make sense and select the variables that we want to be displayed on the table and use the print function to show the data.

     rename(
        per_vax = Series_Complete_Pop_pct_agegroup, 
        date = Date
      ) %>%
      select(date, age_group, per_vax) %>%
      print()

## Results

```{r vax, echo=FALSE}
plot(vax_line_graph)
```

Figure 1. The percentage of different age groups that are fully vaccinated ranging from January 2021 to January 2022.

```{r cases, echo=FALSE}
plot(case_graph)
```

Figure 2. The number of total positive COVID-19 cases from December 2021 to January 2022 in different age groups.

## Discussion

In the case data, the most drastic difference in number of cases was in the 21-30 year old age group. This is similar to the vaccination data where the age group of 18-24 year old was the lowest, disregarding the 5-11 age group because that group was only recently approved to be fully vaccinated . This subgroup of young adults is particularly swayed by herd immunity, and looking into social norms, showed "relatively poor adherence to physical distancing guidelines" (Graupensperger, et al.).

Historical evidence from influenza pandemics show "that pandemics tend to come in waves over the first 2--5 years as the population immunity builds-up (naturally or through vaccination), and then the number of infected cases tends to decrease" (Petersen, et al.). Here we are into year two of the COVID-19 pandemic, and while it seems that the worst is over, there is still so much that could happen in the next few years as far as vaccine development, testing, epidemiological studies, etc. However, as this pandemic continues, and will eventually come to an end, studies like this could be useful for future research on COVID-19.

## References

Graupensperger, Scott, et al. "Social Norms and Vaccine Uptake: College Students' COVID Vaccination Intentions, Attitudes, and Estimated Peer Norms and Comparisons with Influenza Vaccine." ScienceDirect, 8 Apr. 2021, https://www-sciencedirect-com.trmproxy.mnpals.net/science/article/pii/S0264410X21002863.

Hadley Wickham and Jennifer Bryan (2019). readxl: Read Excel Files. R package version 1.3.1. https://CRAN.R-project.org/package=readxl.

Nguyen, Kimberly H, et al. "COVID-19 Vaccination Intent, Perceptions, and Reasons for Not Vaccinating among Groups Prioritized for Early Vaccination --- United States, September and December 2020." Wiley Online Library, 31 Mar. 2021, https://onlinelibrary.wiley.com/doi/full/10.1111/ajt.16560. 

Petersen, Eskild, et al. "Comparing SARS-CoV-2 with SARS-CoV and Influenza Pandemics." The Lancet Infectious Diseases, vol. 20, no. 9, 2020, pp. e238-e244. ProQuest, https://trmproxy.mnpals.net/login?url=https://www.proquest.com/scholarly-journals/comparing-sars-cov-2-with-influenza-pandemics/docview/2437381519/se-2?accountid=12548, doi:http://dx.doi.org/10.1016/S1473-3099(20)30484-9.

Pogue, Kendall et al. "Influences on Attitudes Regarding Potential COVID-19 Vaccination in the United States." Vaccines vol. 8,4 582. 3 Oct. 2020, doi:10.3390/vaccines8040582.

RStudio Team (2020). RStudio: Integrated Development for R. RStudio, PBC, Boston, MA URL http://www.rstudio.com/.

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686.
